From physical principles to ???

### 2016-01-13 2015CompleteDerivation/completeDerivation.tex
Introduction and detrev reviewed

Classical particle mechanics is usually founded on Newton's laws. These, though, are insufficient to derive the full Lagrangian and Hamiltonian formalism, and usually other ad-hoc assumptions (e.g. conservative forces) are introduced. Special relativity is based on two principles (invariance of the speed of light and the principle of relativity), which lead to the Minkowskian nature of space-time but not to the equations of motion, which are a consistent reformulation of the non-relativistic ones. Quantum mechanics is simply founded on its mathematical formulation, leaving its physical significance open to debate. One can't come to the conclusion that, unfortunately, the landscape of fundamental physics is a piece-meal of concepts and idea, sometime apparently inconsistent with each other. Yet: the physical world is one, and our physical theories are simply models of such reality. If we clarify what those models represent, and on what assumptions are based, shouldn't we be able to have a coherent and more satisfying picture?

This work aims to re-organize the known elements and equations in a more consistent and comprehensive way, leading to better insight on why the fundamental concepts and laws are what they are. We will start with physical assumptions, that clarify the models we impose on the physical world, and give arguments on when those assumptions can be considered valid. We will justify our mathematical definitions, based on the formal properties of the objects of our discussion. And will, as a consequence, re-derive known results and theories. We will strive to do this in a way that is mathematically meaningful, philosophically consistent and mathematically precise.

@@

@@

By deterministic (and reversible) we mean that future (and past) states are uniquely identified by the present state. We can then think of the state of the system as the part that interacts deterministically and reversibly with the environment (and itself) while the unstated part as interacting non-deterministically with the environment (and itself). We call this setup the \emph{fundamental model of physics}, as it is the simplest model that still captures the crucial aspects of studying and writing laws for a physical system.

Much of the focus later will be given to the state under deterministic and reversible evolution. However, the non-deterministic part plays a fundamental role as well. In fact: it is this aspect of the evolution that determines what states are available and their description. Suppose we study the motion of a cannonball, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules scatter off its surface randomly, its trajectory is not greatly affect as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory, and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the set of states must be closed under the non-deterministic evolution as well. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, the state gets knocked out of the set of states determined by a simple position/momentum pair.

@@

As we have seen, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the individual trajectories are not (those fall within the unstated part). In fact: we cannot assume trajectories and states are always defined at all for the unstated part. Consider a muon and its decay into an electron and two neutrinos: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal state and trajectory to those parts. The state of a muon is not some combination of the state of an electron and two neutrinos. That is: the unstated part is not just microstates.

@@

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and system theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can be indeed calculated, measured and prepared, not each individual element.

So, if it seems we need to require deterministic and reversible motion, why we call this an assumption? First of all, because it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it studied through experimentation?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they little affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the states consists of the position and momentum of the center of mass means assuming that the Brownian motion of the body is negligible.

@@

\begin{justification}
The collection of all possible states forms a set.
\end{justification}

>>>>

While some branches of fundamental physics (e.g. Newtonian mechanics and special relativity) are founded on physical laws or principles others (e.g. Lagrangian/Hamiltonian/Quantum mechanics) start by postulating mathematical frameworks or relationships. The latter approach presents a significant drawback: the mathematical structure is not enough to define the physical system it describes. We are left asking: under what physical conditions a particular system is described by Lagrangian (or Quantum) mechanics? Naturally one \emph{can} define a Lagrangian system (or a Quantum system) as one that obeys those rules, but that begs the question. Another issue is that each element of the mathematical structure may or may not correspond to some physical concept. Suppose the earning (or losses) of two companies are governed by Hamiltonian's equations, which variable is on the cotangent space of the other? In other words: if we want to fully appreciate what the fundamental theories of physics are describing, what each mathematical concept corresponds to in the physical world, we need to start by characterizing the physical system we are describing in a precise enough way that it can be encoded in mathematical language. Naturally, the physical insight we gain will be relevant for only that type of system (i.e. other different systems may \emph{happen} to use the same equations), but what we'll see is that the assumptions we make are minimal and can apply to a variety of systems, especially the ones that are considered fundamental. It is indeed surprising how so much can be derived by imposing so little.

@@

%This work aims to re-organize the known elements and equations in a more consistent and comprehensive way, leading to better insight on why the fundamental concepts and laws are what they are. We will start with physical assumptions, that clarify the models we impose on the physical world, and give arguments on when those assumptions can be considered valid. We will justify our mathematical definitions, based on the formal properties of the objects of our discussion. And will, as a consequence, re-derive known results and theories. We will strive to do this in a way that is mathematically meaningful, philosophically consistent and mathematically precise.

@@

As the terms used have different meaning in different contexts, let's first clarify them. By deterministic (and reversible) we mean that future (and past) states are uniquely identified by the present state. We can then think of the state of the system as the part that interacts deterministically and reversibly with the environment (and itself) while the unstated part as interacting non-deterministically with the environment (and itself). We call this setup the \emph{fundamental model of physics}, as it is the simplest model that still captures the crucial aspects of studying and writing laws for a physical system.

Much of the focus later will be given to the state under deterministic and reversible evolution. However, the non-deterministic part plays a fundamental role as well. In fact: it is this aspect of the evolution that determines what states are available and their description. Suppose we study the motion of a cannonball, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules scatter off its surface randomly, its trajectory is not greatly affect as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the set of states must be closed under the non-deterministic evolution as well. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, the state gets knocked out of the set of states determined by a simple position/momentum pair.

@@

As we have seen, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the trajectories of the parts are not (those fall within the unstated part). In fact: we \emph{cannot} assume trajectories and states are always defined at all for the unstated part. Consider a muon and its decay into an electron and two neutrinos: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal independent state and evolution to each part. The state of a muon is not some combination of the state of an electron and two neutrinos. That is: the unstated part is not just microstates.

@@

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and system theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can be indeed calculated, measured and prepared. We can also describe the evolution of each element of the distribution provided that: we have a way to isolate it and study it under deterministic and reversible motion (so that we can define its states); the non-deterministic motion does not alter the system (the set of states remains the same). For example, consider a wooden ball described by position and momentum of the center of mass: if randomly scattered by other balls retain the set of states, if thrown into a fire it does not.

So, if it seems we need to require deterministic and reversible motion, why we call this an assumption? First of all, because it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it studied through experimentation?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they little affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the states consists of the position and momentum of the center of mass requires assuming that the Brownian motion of the body is negligible.

@@

\begin{justification}
The collection of all possible states forms a set as we can physically distinguish between them.
\end{justification}
